Title

### 2012-11-20 inc/laser/laser.tex
move more of apl paper to este
added several of the opt express graphics

\section{Yb:KGW Laser System}
\subsection{Introduction}
% Draws from:
% High-power, femtosecond, thermal-lens-shaped Yb:KGW oscillator
% Joel A. Berger, Michael J. Greco, and W. Andreas Schroeder
% 9 June 2008 / Vol. 16, No. 12 / OPTICS EXPRESS

@@

The design of the laser head follows the example of Holtom \cite{Holtom_mode_locked_2006}, which ensures efficient pumping and access to the broadest emission bandwidth of Yb:KGW, while also employing the thermal lens shaping (TLS) technique to compensate for astigmatism \cite{Rimington_thermal_lens_2004}.
To ensure efficient operation of a Yb:KGW laser, the dominant absorption feature at 980nm for radiation polarized parallel to the optical $N_m$-axis ($\sim\,20^\circ$ from the crystallographic a-axis) should be pumped using high brightness diode lasers.
On the other hand, as with \ce{Yb$:$KY(WO4)2} \cite{Liu_diode_pumped_2001,Killi_high_peak_2005}, the broadest emission bandwidth for ultrashort pulse generation occurs for radiation polarized parallel to the optical $N_p$-axis (parallel to the crystallographic $b$-axis \cite{Holtom_mode_locked_2006}) in this biaxial material. 
%page 8632
Instead of using a polarization-coupled pumping geometry \cite{Holtom_mode_locked_2006} to ensure these optimum pump-lasing conditions, we employ the non-Brewster crystal geometry and cut shown in Fig. 1(a), where a single 193nm-thick \ce{SiO2} anti-reflection coating is applied to both polished 3x10mm Yb:KGW crystal faces.
At an angle of incidence of 32.3$^\circ$, this coating theoretically has zero reflection loss for $p$-polarized 1040nm laser radiation parallel to the $b$-axis propagating perpendicular to the $a$-$b$ plane of the crystal (Fig. 1(b)); i.e., the preferred optical $N_p$-axis with refractive index $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}.
The \ce{SiO2} coating also has an appropriate bandwidth for femtosecond laser operation in excess of 100nm (to the
0.1\% reflection points) for $p$-polarized radiation centered at 1040nm.
On the other hand, the $\sim$ 0.4\% reflection loss per surface for $s$-polarized 1040nm radiation is sufficient to suppress $s$-polarized oscillation in a laser resonator.
In addition, the same anti-reflection coating has less than 0.5\% reflection loss for $s$-polarized 980nm pump radiation (polarized parallel to the $a$-axis of the Yb:KGW crystal) at incidence angles between 21 and 39$^\circ$ (Fig. 1(b)).
Thus, in this Yb:KGW laser crystal geometry, the largest absorption cross-section associated with the optical $N_m$-axis is accessed with about 94\% efficiency ($\cos 20^\circ = 0.94$) by the $s$-polarized pump radiation.

\begin{figure}
  \subfloat[][]{
    \label{fig:crystal_geo}
    \input{crystal_geo}
  }
  \caption{
\protect\subref{fig:crystal_geo} Top view of the 3x3x10mm AR coated Yb:KGW laser crystal showing the crystal cut, the direction of propagation of p-polarized laser radiation at 1040nm, and the $x$ (tangential) and $y$ (saggital) directions (in blue) employed in the thermal analysis;
(b) Calculated reflectivity of the 193nm \ce{SiO2} AR coating as a function of incidence angle for $p$-polarized 1040nm laser radiation (solid line) and $s$-polarized 980nm pump radiation (dashed line).
}
\end{figure}

@@

To ensure a stigmatic laser resonator, we also employ the thermal lens shaping (TLS) technique \cite{Rimington_thermal_lens_2004} to compensate for the astigmatism induced by the non-normal incidence geometry of the laser crystal. 
For the employed non-Brewster crystal cut (Fig. 1(a)), a ray
%page 8633
transfer matrix analysis (with the $x$ direction parallel to the crystallographic $b$-axis and the $y$
direction parallel to the $a$-axis) reveals that the required ellipticity for a stigmatic thermal lens
generated by a parabolic temperature distribution, $T(x, y) = T_0 - \frac{1}{2} ( A_\smallT x^2 + B_\smallT y^2 )$, where $T_0$
is the on-axis temperature, is given by
%TODO check W symbol consistent with the rest of the paper
\begin{equation} \label{eq:laser_ratio_ba}
  \frac{B_\smallT}{A_\smallT} = 
  \frac{ \cos^2 \theta_2 \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) }
       { \cos^2 \theta_1 \left[ \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) \right] }
  \quad\text{.}
\end{equation}
Here, $\frac{dn}{dT}$ and $n$ are the thermo-optic coefficient and refractive index of the medium respectively (in this case for radiation polarized along the optical $N_p$-axis), $\theta_1$ is the 32.3$^\circ$ angle of incidence, $\theta_2$ = 15.7$^\circ$ is the angle of refraction (as $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}), and $\Delta$ is the fraction of the crystal length $l$ contributing to the bowing of the crystal faces under thermal
expansion with coefficient $\alpha T$ \cite{Rimington_thermal_lens_2004}.
Generally, $\Delta \approx W l$ , where $W$ is the transverse size (radius) of the pumped region \cite{Koechner_thermal_1970}.
In our case, $l \approx$ 3mm and $W \approx$ 0.2mm, so that $\Delta < 0.1$, implying that the thermal duct due to a non-zero $\frac{dn}{dT}$ primarily determines the required $B_T$/$A_T$ ratio.
In other words, for the potassium gadolinium tungstate (\ce{KGd(WO4)2}) crystal geometry of Fig. 1(a), we expect from \ref{eq:laser_ratio_ba} to require $\dfrac{B_\smallT}{A_\smallT} \approx \dfrac{ \cos^2 \theta_2 }{ \cos^2 \theta_1 } = 1.3$.

The spatial pump distribution needed to generate the required $B_T/A_T$ ratio can be found using a cosine Fourier series solution to the thermal diffusion equation for our two-dimensional problem,
\begin{equation} \label{eq:heat_eqn}
  \kappa \nabla^2 T(x,y) = Q(x,y)
  \quad\text{,}
\end{equation}
where $\kappa$ is the thermal diffusion coefficient.
Thermal diffusion in the third $z$-direction has been neglected due to the relatively uniform longitudinal heat deposition resulting from our counter-propagating pump geometry \cite{Rimington_thermal_lens_2004}.
The specifications of our dual-axis collimated diodes \cite{website_limo} indicate that heat source can be written as
\begin{equation}
  Q(x,y) = Q_0 \exp \left[ - \frac{x^2}{W_x^2} - \left( \frac{y^2}{W_y^2} \right)^3 \right]
  \quad\text{;}
\end{equation}
that is, the focused pump radiation is described accurately by a Gaussian irradiance distribution in the horizontal ($x$) direction with half-width 1/e maximum (HW1/eM) spot size $W_x$ and a super-Gaussian of order 3 in the vertical ($y$) direction with spot size $W_y$.
\ref{eq:heat_eqn} can then be solved using the appropriate boundary conditions for our 3x10mm crystal cross-section for the employed TLS technique \cite{Rimington_thermal_lens_2004}; namely, heat removal only through the top crystal face implies $T(x,y = \pm 1.5\text{mm}) = 0$ (or arbitrary constant) and $\dfrac{\partial T}{\partial x} = 0$ at $ x = \pm 5.0$mm for any $y$.
For a fixed vertical pump spot size of $W_y = 0.2$mm, our analysis indicates that in order to achieve the required ratio $B_T / A_T = 1.3$ in the parabolic approximation for $T(x,y)$ about $(x,y) = (0,0)$ a horizontal pump spot size $W_x = 0.3$mm is needed; in other words, the required $Q(x,y)$ heat source spot size ratio (or ellipticity) $W_x / W_y = 1.5$.
This analytical result for the central region around the laser beam axis is shown in Fig. 2; displayed are the
%page 8634
horizontal ($y = 0$, Fig. 2(a)) and vertical ($x = 0$, Fig. 2(b)) sections, respectively, through the $Q(x,y)$ and $T(x,y)$ distributions.
Also shown are the parabolic approximations to the pump-induced temperature distribution (dashed lines) and representative TEM$_{00}$ Gaussian laser modes (shaded area) in the Yb:KGW crystal with a HW1/eM field mode sizes of $w_y = 160\mu$m and $w_x = 182 \mu$m (14\% larger due to refraction (Fig. 1(a)).
This mode size ensures efficient pump-probe overlap while minimizing the loss due to ground state absorption in the unpumped regions of the Yb-doped gain medium \cite{Brenier_new_criteria}; in particular, in the vertical direction (Fig. 2(b)) where $w_y$ greater than about $0.8 W_y$ leads to significant absorptive loss as the extrema of the laser mode extend beyond the sharply-bounded super-Gaussian pump region.

% insert figure 2

@@

We note that the above thermal lens analysis neglects the anisotropy of the thermal conductivity $\kappa$ in potassium gadolinium tungstate (\ce{KGd(WO4)2}) \cite{Biswal_thermo_optical_05,mochalov_laser_1997} and stress-induced refractive index changes \cite{Yumashev_laser_2007}.
For the employed crystal cut in our TLS geometry (Fig. 1(a)), the anisotropy in $\kappa$ \cite{mochalov_laser_1997,website_EKSPLA} is not expected to have a significant effect; the laser head geometry dictates that the dominant pump-induced heat conduction will be in the saggital ($y$) direction along the crystal $a$-axis where $\kappa \approx 2.6$W/(mK) (heat extraction only through the top and bottom 3x10mm crystal faces), so that the 46\% larger value of $\kappa$ along the $b$-axis ($x$) direction at 3.8W/(mK) should not have a large influence on the heat conduction.
The refractive index
%page 8635
changes due to stress in the pumped gain medium are difficult to quantify since the substantial variations in the reported values of the thermo-optic coefficients in KGW make it impossible to extract either the sign or magnitude of the stress dependence of the refractive index \cite{website_time_bandwidth}.
Moreover, nearly all analyses of thermal lensing effects Nd:KGW and Yb:KGW lasers have been performed for laser radiation polarized along the optical $N_m$-axis, which accesses the largest emission cross-section for both Nd- and Yb-doping \cite{Yumashev_laser_2007,Hellstrom_efficient_2006}.
Consequently, for our laser polarized along the optical $N_p$-axis \cite{Holtom_mode_locked_2006}, there is insufficient information to include accurately the effects of stress-induced refractive index changes.

@@

The diode-pumped TLS Yb:KGW laser head is placed in the 2.35m-long, asymmetric z-fold cavity shown in Fig. 3(a).
The two focusing gain section mirrors each have a group velocity dispersion (GVD) of $-1300 ( \pm 150 ) \text{fs}^2$ (Layertec GmbH \cite{website_layertec}), a radius of curvature $R$ = 50cm, and are positioned 29.5cm from the gain medium.
The 5$^\circ$ angle of incidence on the gain section mirrors results in minimal astigmatism, although any net round-trip cavity astigmatism can be compensated for by the TLS technique.
The longer 75cm arm of the resonator is terminated by a concave ($R$ = 1m and angle of incidence $< 2^\circ$) high reflector

\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:laser_schematic}
    \input{schematic}
  }
  \caption{
\protect\subref{fig:laser_schematic} Schematic of the femtosecond TLS Yb:KGW oscillator (LD = laser diode, OC = output coupler, SBR = saturable Bragg reflector).
(b) Laser mode size in the center of the Yb:KGW gain medium ($w$) as a function of the strength of the lensing ($1/f$) in the gain medium (thermal and Kerr) obtained from a resonator stability analysis with $z \approx 42$cm for both the tangential ($x$, red line) and saggital ($y$, dashed black line) directions.
The 0.2mm (HW1/eM) vertical super-Gaussian pump spot size is indicated by the horizontal line.
}
\end{figure}

@@

The stability analysis of our diode-pumped Yb:KGW laser (Fig. 3(b)), clearly indicates that the resonator configuration is unstable for thermal focal lengths less than about $-6m^{-1}$ (diopters) and is unlikely to oscillate for negative thermal focal lengths as this requires a vertical TEM$_{00}$-mode size $w_y$ greater than 180$\mu$m in the Yb:KGW gain medium --- a value that would result in strong absorptive losses due to the restrictive $W_y = 200\mu$m super-Gaussian vertical pump beam size in the Yb-doped crystal.
The clear implication is therefore that the net thermal lens due to the dominant thermal duct is positive; that is, the sum of the refractive index change due to the thermo-optic coefficient and stress for 1040nm laser radiation polarized along the optical $N_p$-axis (crystal $b$-axis) is positive.
This result is in apparent contradiction to some recent measurements of relatively large negative thermo-optic coefficients in Yb:KGW \cite{Biswal_thermo_optical_05}, but is consistent with other determinations of the pump-induced thermal focal length for laser oscillation polarized parallel to the optical $N_p$-axis \cite{Holtom_mode_locked_2006,Hellstrom_efficient_2006}.

@@

In the cavity configuration depicted in Fig. 3(a), self-starting mode-locked (ML) operation was observed for output powers greater than $\sim$2W and $z \approx 42$cm.
The maximum (or optimum) mode-locked output power obtained without the emergence of DC spectral components was 3.5W, with a sech$^2$ pulse duration of 250fs (Fig. 4(a)).
The insertion of two plane dispersion compensation mirrors each with a GVD of $-700\text{fs}^2$ (Layertec GmbH \cite{website_layertec}) into the short arm of the cavity generated higher mode-locked output powers with longer pulse durations.
For example, with two reflections off each mirror (an additional $-5600\text{fs}^2$ of round-trip GVD), the pulse duration increased to 347fs (Fig. 4(b)) at an optimum mode-locked output power of $5.0$W.
This trend is consistent with the expected `soliton' modelocking regime; that is, the SBR only initiates the mode-locking mechanism, which is quickly dominated by self-phase modulation in the Yb:KGW gain medium balanced by the net negative GVD in the cavity \cite{Brunner_diode_pumped}.

The laser is capable of generating shorter ML pulse durations.
Replacement of one of the $-1300 ( \pm 150 ) \text{fs}^2$ z-fold mirrors with a conventional $R = 50$cm high reflector and the insertion of $-880\text{fs}^2$ of negative GVD per round trip (a pair of $-220\text{fs}^2$ dispersion compensating mirrors) generated $\sim220$fs pulses at 24A of diode current, 2W of output power, and a SBR position $z = 43.2$cm.
However, ML operation was significantly less stable under these conditions, indicating that either the cavity is operating close to the net zero GVD point or that the finite gain bandwidth is influencing the ML operation.
We also note that in this case the optimum SBR position $z$ is larger than that for higher power 3.5W ML operation with a negative cavity GVD of $-5,200\text{fs}^2$ per round-trip (Fig. 4(a)), but is always less than the 50cm focal length of the concave high reflector.
This is due to the need to match the absorption-induced thermal bowing of the SBR (i.e., the radius of curvature of its front surface mirror) to the radius of curvature of the intra-cavity laser mode \cite{Schieffer_dual_passive_2006}.
In general, good ML operation is achieved with the SBR positioned within $\pm$1cm from the optimum $z$ distance.

Figure 5 shows the laser's output power as a function of diode current for the cavity configuration depicted in Fig. 3(a); a total negative GVD of $-5,200\text{fs}^2$ per round-trip and a 7\% transmission output coupler.
Soliton mode-locked (ML) operation is observed for diode currents greater than 23A, corresponding to a continuous-wave (cw) output power of 1.8W.
Between diode currents (output powers) of 23 and 25A (1.8 and 2.3W), the resonator operates mainly in the cw mode, but will transition to ML operation upon being subjected to a perturbation.
The switching to ML operation is accompanied by a 250mW increase in output
%page 8637
power indicating a higher cavity Q for ML operation.
This is consistent both with the laser accessing more of the available inhomogeneous gain bandwidth and with our resonator design favoring operation with a strong Kerr lens in the Yb:KGW gain medium.
Given a value of $n^2 \approx 15 \cdot 10^{-16}\text{cm}^2/\text{W}$ for 1040nm radiation polarized parallel to the KGW crystal's optical $N_p$-axis \cite{Major_characterization_2003,Selivanov_nonlinear_2006,Vodchits_zscan_2006} and a calculated peak pulse irradiance of $\sim$4GW/cm$^2$ in the gain medium at 24A, we expect a Kerr lens strength of approximately 2m$^-1$.
From Fig. 3(b), we see that the resulting increase in the strength of the lensing in the Yb:KGW crystal will result in a reduction of the laser mode size in the gain medium, which produces a better spatial match to gain duct produced by the laser diode pumps (less absorptive loss), thus generating an increase in output power.

% figure 4

>>>>

\section{Yb:KGW Laser System} \label{sec:laser}
\subsection{Introduction}
% Draws from:
% High-power, femtosecond, thermal-lens-shaped Yb:KGW oscillator
% Joel A. Berger, Michael J. Greco, and W. Andreas Schroeder
% 9 June 2008 / Vol. 16, No. 12 / OPTICS EXPRESS

@@

The design of the laser head follows the example of Holtom \cite{Holtom_mode_locked_2006}, which ensures efficient pumping and access to the broadest emission bandwidth of Yb:KGW, while also employing the thermal lens shaping (TLS) technique to compensate for astigmatism \cite{Rimington_thermal_lens_2004}.
To ensure efficient operation of a Yb:KGW laser, the dominant absorption feature at 980nm for radiation polarized parallel to the optical $N_m$-axis ($\sim\,20^\circ$ from the crystallographic a-axis) should be pumped using high brightness diode lasers.
On the other hand, as with \ce{Yb$:$KY(WO4)2} \cite{Liu_diode_pumped_2001,Killi_high_peak_2005}, the broadest emission bandwidth for ultrashort pulse generation occurs for radiation polarized parallel to the optical $N_p$-axis (parallel to the crystallographic $b$-axis \cite{Holtom_mode_locked_2006}) in this biaxial material. 
%page 8632
Instead of using a polarization-coupled pumping geometry \cite{Holtom_mode_locked_2006} to ensure these optimum pump-lasing conditions, we employ the non-Brewster crystal geometry and cut shown in \ref{fig:crystal}\subref{fig:crystal_geo}, where a single 193nm-thick \ce{SiO2} anti-reflection coating is applied to both polished 3x10mm Yb:KGW crystal faces.
At an angle of incidence of 32.3$^\circ$, this coating theoretically has zero reflection loss for $p$-polarized 1040nm laser radiation parallel to the $b$-axis propagating perpendicular to the $a$-$b$ plane of the crystal (\ref{fig:crystal}\subref{fig:crystal_coating}); i.e., the preferred optical $N_p$-axis with refractive index $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}.
The \ce{SiO2} coating also has an appropriate bandwidth for femtosecond laser operation in excess of 100nm (to the
0.1\% reflection points) for $p$-polarized radiation centered at 1040nm.
On the other hand, the $\sim$ 0.4\% reflection loss per surface for $s$-polarized 1040nm radiation is sufficient to suppress $s$-polarized oscillation in a laser resonator.
In addition, the same anti-reflection coating has less than 0.5\% reflection loss for $s$-polarized 980nm pump radiation (polarized parallel to the $a$-axis of the Yb:KGW crystal) at incidence angles between 21 and 39$^\circ$ (\ref{fig:crystal}\subref{fig:crystal_coating}).
Thus, in this Yb:KGW laser crystal geometry, the largest absorption cross-section associated with the optical $N_m$-axis is accessed with about 94\% efficiency ($\cos 20^\circ = 0.94$) by the $s$-polarized pump radiation.

\begin{figure}
  \subfloat[][]{
    \label{fig:crystal_geo}
    \input{crystal_geo}
  }
  \subfloat[][]{
    \label{fig:crystal_coating}
    \includegraphics{coating.jpg}
  }
  \caption{
    \protect\subref{fig:crystal_geo} Top view of the 3x3x10mm AR coated Yb:KGW laser crystal showing the crystal cut, the direction of propagation of p-polarized laser radiation at 1040nm, and the $x$ (tangential) and $y$ (saggital) directions (in blue) employed in the thermal analysis;
    \protect\subref{fig:crystal_coating} Calculated reflectivity of the 193nm \ce{SiO2} AR coating as a function of incidence angle for $p$-polarized 1040nm laser radiation (solid line) and $s$-polarized 980nm pump radiation (dashed line).
  }
  \label{fig:crystal}
\end{figure}

@@

To ensure a stigmatic laser resonator, we also employ the thermal lens shaping (TLS) technique \cite{Rimington_thermal_lens_2004} to compensate for the astigmatism induced by the non-normal incidence geometry of the laser crystal. 
For the employed non-Brewster crystal cut (\ref{fig:crystal}\subref{fig:crystal_geo}), a ray
%page 8633
transfer matrix analysis (with the $x$ direction parallel to the crystallographic $b$-axis and the $y$
direction parallel to the $a$-axis) reveals that the required ellipticity for a stigmatic thermal lens
generated by a parabolic temperature distribution, $T(x, y) = T_0 - \frac{1}{2} ( A_\smallT x^2 + B_\smallT y^2 )$, where $T_0$
is the on-axis temperature, is given by
%TODO check W symbol consistent with the rest of the paper
\begin{equation} \label{eq:laser_ratio_ba}
  \frac{B_\smallT}{A_\smallT} = 
  \frac{ \cos^2 \theta_2 \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) }
       { \cos^2 \theta_1 \left[ \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) \right] }
  \quad\text{.}
\end{equation}
Here, $\frac{dn}{dT}$ and $n$ are the thermo-optic coefficient and refractive index of the medium respectively (in this case for radiation polarized along the optical $N_p$-axis), $\theta_1$ is the 32.3$^\circ$ angle of incidence, $\theta_2$ = 15.7$^\circ$ is the angle of refraction (as $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}), and $\Delta$ is the fraction of the crystal length $l$ contributing to the bowing of the crystal faces under thermal
expansion with coefficient $\alpha T$ \cite{Rimington_thermal_lens_2004}.
Generally, $\Delta \approx W l$ , where $W$ is the transverse size (radius) of the pumped region \cite{Koechner_thermal_1970}.
In our case, $l \approx$ 3mm and $W \approx$ 0.2mm, so that $\Delta < 0.1$, implying that the thermal duct due to a non-zero $\frac{dn}{dT}$ primarily determines the required $B_T$/$A_T$ ratio.
In other words, for the potassium gadolinium tungstate (\ce{KGd(WO4)2}) crystal geometry of \ref{fig:crystal}\subref{fig:crystal_geo}, we expect from \ref{eq:laser_ratio_ba} to require $\dfrac{B_\smallT}{A_\smallT} \approx \dfrac{ \cos^2 \theta_2 }{ \cos^2 \theta_1 } = 1.3$.

The spatial pump distribution needed to generate the required $B_T/A_T$ ratio can be found using a cosine Fourier series solution to the thermal diffusion equation for our two-dimensional problem,
\begin{equation} \label{eq:heat_eqn}
  \kappa \nabla^2 T(x,y) = Q(x,y)
  \quad\text{,}
\end{equation}
where $\kappa$ is the thermal diffusion coefficient.
Thermal diffusion in the third $z$-direction has been neglected due to the relatively uniform longitudinal heat deposition resulting from our counter-propagating pump geometry \cite{Rimington_thermal_lens_2004}.
The specifications of our dual-axis collimated diodes \cite{website_limo} indicate that heat source can be written as
\begin{equation}
  Q(x,y) = Q_0 \exp \left[ - \frac{x^2}{W_x^2} - \left( \frac{y^2}{W_y^2} \right)^3 \right]
  \quad\text{;}
\end{equation}
that is, the focused pump radiation is described accurately by a Gaussian irradiance distribution in the horizontal ($x$) direction with half-width 1/e maximum (HW1/eM) spot size $W_x$ and a super-Gaussian of order 3 in the vertical ($y$) direction with spot size $W_y$.
\ref{eq:heat_eqn} can then be solved using the appropriate boundary conditions for our 3x10mm crystal cross-section for the employed TLS technique \cite{Rimington_thermal_lens_2004}; namely, heat removal only through the top crystal face implies $T(x,y = \pm 1.5\text{mm}) = 0$ (or arbitrary constant) and $\dfrac{\partial T}{\partial x} = 0$ at $ x = \pm 5.0$mm for any $y$.
For a fixed vertical pump spot size of $W_y = 0.2$mm, our analysis indicates that in order to achieve the required ratio $B_T / A_T = 1.3$ in the parabolic approximation for $T(x,y)$ about $(x,y) = (0,0)$ a horizontal pump spot size $W_x = 0.3$mm is needed; in other words, the required $Q(x,y)$ heat source spot size ratio (or ellipticity) $W_x / W_y = 1.5$.
This analytical result for the central region around the laser beam axis is shown in \ref{fig:laser-thermal}; displayed are the
%page 8634
horizontal ($y = 0$, \ref{fig:laser-thermal}(a)) and vertical ($x = 0$, \ref{fig:laser-thermal}(b)) sections, respectively, through the $Q(x,y)$ and $T(x,y)$ distributions.
Also shown are the parabolic approximations to the pump-induced temperature distribution (dashed lines) and representative TEM$_{00}$ Gaussian laser modes (shaded area) in the Yb:KGW crystal with a HW1/eM field mode sizes of $w_y = 160\mu$m and $w_x = 182 \mu$m (14\% larger due to refraction (\ref{fig:crystal}\subref{fig:crystal_geo}).
This mode size ensures efficient pump-probe overlap while minimizing the loss due to ground state absorption in the unpumped regions of the Yb-doped gain medium \cite{Brenier_new_criteria}; in particular, in the vertical direction (\ref{fig:laser-thermal}(b)) where $w_y$ greater than about $0.8 W_y$ leads to significant absorptive loss as the extrema of the laser mode extend beyond the sharply-bounded super-Gaussian pump region.

\begin{figure}
  \centering
  \includegraphics{thermal.jpg}
  \caption{
    Analytical evaluation of the temperature distribution $T(x,y)$ due to the Gaussian/super-Gaussian heat source $Q(x,y)$ in the (a) horizontal ($y=0$) and (b) vertical ($x=0$) directions: $Q(x,y)$ - red-brown line; $T(x,y)$ - black line; parabolic approximation to $T(x,y)$ - dashed black line; representative TEM$_{00}$ laser field modes - solid green.
  }
  \label{fig:laser-thermal}
\end{figure}

@@

We note that the above thermal lens analysis neglects the anisotropy of the thermal conductivity $\kappa$ in potassium gadolinium tungstate (\ce{KGd(WO4)2}) \cite{Biswal_thermo_optical_05,mochalov_laser_1997} and stress-induced refractive index changes \cite{Yumashev_laser_2007}.
For the employed crystal cut in our TLS geometry (\ref{fig:crystal}\subref{fig:crystal_geo}), the anisotropy in $\kappa$ \cite{mochalov_laser_1997,website_EKSPLA} is not expected to have a significant effect; the laser head geometry dictates that the dominant pump-induced heat conduction will be in the saggital ($y$) direction along the crystal $a$-axis where $\kappa \approx 2.6$W/(mK) (heat extraction only through the top and bottom 3x10mm crystal faces), so that the 46\% larger value of $\kappa$ along the $b$-axis ($x$) direction at 3.8W/(mK) should not have a large influence on the heat conduction.
The refractive index
%page 8635
changes due to stress in the pumped gain medium are difficult to quantify since the substantial variations in the reported values of the thermo-optic coefficients in KGW make it impossible to extract either the sign or magnitude of the stress dependence of the refractive index \cite{website_time_bandwidth}.
Moreover, nearly all analyses of thermal lensing effects Nd:KGW and Yb:KGW lasers have been performed for laser radiation polarized along the optical $N_m$-axis, which accesses the largest emission cross-section for both Nd- and Yb-doping \cite{Yumashev_laser_2007,Hellstrom_efficient_2006}.
Consequently, for our laser polarized along the optical $N_p$-axis \cite{Holtom_mode_locked_2006}, there is insufficient information to include accurately the effects of stress-induced refractive index changes.

@@

The diode-pumped TLS Yb:KGW laser head is placed in the 2.35m-long, asymmetric z-fold cavity shown in \ref{fig:laser-cavity}\subref{fig:laser_schematic}.
The two focusing gain section mirrors each have a group velocity dispersion (GVD) of $-1300 ( \pm 150 ) \text{fs}^2$ (Layertec GmbH \cite{website_layertec}), a radius of curvature $R$ = 50cm, and are positioned 29.5cm from the gain medium.
The 5$^\circ$ angle of incidence on the gain section mirrors results in minimal astigmatism, although any net round-trip cavity astigmatism can be compensated for by the TLS technique.
The longer 75cm arm of the resonator is terminated by a concave ($R$ = 1m and angle of incidence $< 2^\circ$) high reflector

\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:laser_schematic}
    \input{schematic}
  }
  \\
  \subfloat[][]{
    \label{fig:laser_stability}
    \includegraphics{stability.jpg}
  }
  \caption{
    \protect\subref{fig:laser_schematic} Schematic of the femtosecond TLS Yb:KGW oscillator (LD = laser diode, OC = output coupler, SBR = saturable Bragg reflector).
    \protect\subref{fig:laser_stability} Laser mode size in the center of the Yb:KGW gain medium ($w$) as a function of the strength of the lensing ($1/f$) in the gain medium (thermal and Kerr) obtained from a resonator stability analysis with $z \approx 42$cm for both the tangential ($x$, red line) and saggital ($y$, dashed black line) directions.
    The 0.2mm (HW1/eM) vertical super-Gaussian pump spot size is indicated by the horizontal line.
  }
  \label{fig:laser-cavity}
\end{figure}

@@

The stability analysis of our diode-pumped Yb:KGW laser (\ref{fig:laser-cavity}\subref{fig:laser_schematic}), clearly indicates that the resonator configuration is unstable for thermal focal lengths less than about $-6m^{-1}$ (diopters) and is unlikely to oscillate for negative thermal focal lengths as this requires a vertical TEM$_{00}$-mode size $w_y$ greater than 180$\mu$m in the Yb:KGW gain medium --- a value that would result in strong absorptive losses due to the restrictive $W_y = 200\mu$m super-Gaussian vertical pump beam size in the Yb-doped crystal.
The clear implication is therefore that the net thermal lens due to the dominant thermal duct is positive; that is, the sum of the refractive index change due to the thermo-optic coefficient and stress for 1040nm laser radiation polarized along the optical $N_p$-axis (crystal $b$-axis) is positive.
This result is in apparent contradiction to some recent measurements of relatively large negative thermo-optic coefficients in Yb:KGW \cite{Biswal_thermo_optical_05}, but is consistent with other determinations of the pump-induced thermal focal length for laser oscillation polarized parallel to the optical $N_p$-axis \cite{Holtom_mode_locked_2006,Hellstrom_efficient_2006}.

@@

In the cavity configuration depicted in \ref{fig:laser-cavity}\subref{fig:laser_schematic}, self-starting mode-locked (ML) operation was observed for output powers greater than $\sim$2W and $z \approx 42$cm.
The maximum (or optimum) mode-locked output power obtained without the emergence of DC spectral components was 3.5W, with a sech$^2$ pulse duration of 250fs (\ref{fig:laser-profile}(a)).
The insertion of two plane dispersion compensation mirrors each with a GVD of $-700\text{fs}^2$ (Layertec GmbH \cite{website_layertec}) into the short arm of the cavity generated higher mode-locked output powers with longer pulse durations.
For example, with two reflections off each mirror (an additional $-5600\text{fs}^2$ of round-trip GVD), the pulse duration increased to 347fs (\ref{fig:laser-profile}(b)) at an optimum mode-locked output power of $5.0$W.
This trend is consistent with the expected `soliton' modelocking regime; that is, the SBR only initiates the mode-locking mechanism, which is quickly dominated by self-phase modulation in the Yb:KGW gain medium balanced by the net negative GVD in the cavity \cite{Brunner_diode_pumped}.

The laser is capable of generating shorter ML pulse durations.
Replacement of one of the $-1300 ( \pm 150 ) \text{fs}^2$ z-fold mirrors with a conventional $R = 50$cm high reflector and the insertion of $-880\text{fs}^2$ of negative GVD per round trip (a pair of $-220\text{fs}^2$ dispersion compensating mirrors) generated $\sim220$fs pulses at 24A of diode current, 2W of output power, and a SBR position $z = 43.2$cm.
However, ML operation was significantly less stable under these conditions, indicating that either the cavity is operating close to the net zero GVD point or that the finite gain bandwidth is influencing the ML operation.
We also note that in this case the optimum SBR position $z$ is larger than that for higher power 3.5W ML operation with a negative cavity GVD of $-5,200\text{fs}^2$ per round-trip \ref{fig:laser-profile}(a)), but is always less than the 50cm focal length of the concave high reflector.
This is due to the need to match the absorption-induced thermal bowing of the SBR (i.e., the radius of curvature of its front surface mirror) to the radius of curvature of the intra-cavity laser mode \cite{Schieffer_dual_passive_2006}.
In general, good ML operation is achieved with the SBR positioned within $\pm$1cm from the optimum $z$ distance.

\ref{fig:laser-performance} shows the laser's output power as a function of diode current for the cavity configuration depicted in \ref{fig:laser-cavity}\subref{fig:laser_schematic}; a total negative GVD of $-5,200\text{fs}^2$ per round-trip and a 7\% transmission output coupler.
Soliton mode-locked (ML) operation is observed for diode currents greater than 23A, corresponding to a continuous-wave (cw) output power of 1.8W.
Between diode currents (output powers) of 23 and 25A (1.8 and 2.3W), the resonator operates mainly in the cw mode, but will transition to ML operation upon being subjected to a perturbation.
The switching to ML operation is accompanied by a 250mW increase in output
%page 8637
power indicating a higher cavity Q for ML operation.
This is consistent both with the laser accessing more of the available inhomogeneous gain bandwidth and with our resonator design favoring operation with a strong Kerr lens in the Yb:KGW gain medium.
Given a value of $n^2 \approx 15 \cdot 10^{-16}\text{cm}^2/\text{W}$ for 1040nm radiation polarized parallel to the KGW crystal's optical $N_p$-axis \cite{Major_characterization_2003,Selivanov_nonlinear_2006,Vodchits_zscan_2006} and a calculated peak pulse irradiance of $\sim$4GW/cm$^2$ in the gain medium at 24A, we expect a Kerr lens strength of approximately 2m$^-1$.
From \ref{fig:laser-cavity}\subref{fig:laser_stability}, we see that the resulting increase in the strength of the lensing in the Yb:KGW crystal will result in a reduction of the laser mode size in the gain medium, which produces a better spatial match to gain duct produced by the laser diode pumps (less absorptive loss), thus generating an increase in output power.

\begin{figure}
  \centering
  \includegraphics{profile.jpg}
  \caption{
Second harmonic autocorrelation measurements and output spectra of the ML Yb:KGW laser; (a) the 250fs laser pulse duration (and fit assuming a sech$^2$ pulse shape) for the cavity depicted in \protect\ref{fig:laser-cavity}\protect\subref{fig:laser_schematic} at an output power of 3.5W, (b) the 347fs pulse duration (and fit assuming a sech$^2$ pulse shape) obtained with an additional -5600fs$^2$ of round-trip GVD at an output power of 5W, (c) the 4.7nm full-width half maximum spectrum associated with the 250fs ML pulse of (a), and (d) spectral output for the ML resonator of \protect\ref{fig:laser-cavity}\protect\subref{fig:laser_schematic} at a laser power of 3.7W.
  }
  \label{fig:laser-profile}
\end{figure}
